Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-10-25 manuscript/manuscript.Rmd
update discussion a bit

The capacity of predators to match their tactic to the behaviour of their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills and expertise to successfully hunt their prey, which may result in individual foraging specialization. Yet, there are very limited empirical assessments showing links between experience, foraging specialization, and hunting success at the individual level, due to the numerous challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to investigate if and how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. Our results show that predators optimized prey consumption as they gained experience. Prey behaviour was an important mediator of this relationship, resulting in differences among predators in the development of expertise. Throughout experience, 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by swithing between slow and fast speeds. The remaining 32% of predators used a specialized tactic of fast movement throughout the study period. Contrary to our predictions, the predator's strategies (i.e. flexible or specialized foraging) was not strongly matched to the preys' speed, suggesting that changes in hunting behaviour were driven more by repeated experience than the prey's speed. Ultimately, flexible foragers were only slightly less successful than specialist foragers, such that both achieved similar success overall. Our observations suggest that experience drives among individual differences in hunting tactics, which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [Refs]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are unrealistic for animals in nature [refs]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@FraserFranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. In a different system, @lymberyetal.2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

@@

Learning from experience is predicted to increase individual predator foraging specialization when prey distribution, behaviour, and movement is predictable through time [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, if predators encounter varying prey with a greater behavioural range through time, then learning should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022]. We found that predators in *DBD* increased their hunting success with experience, suggesting that the development of predator expertise was honed through extensive practice. While there was no directional increase in either behavioural specialization or flexibility with experience at the population level, individual predators developed their own behavioural trajectories over successive encounters with the prey. This was characterized by a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy, with some predators switching between the two with experience. The behavioural strategies of predators and prey tended to match across experience, suggesting that both were probably adjusting to each other. Specialized cursorial hunters tended to encounter similar groups of prey, while slower and flexible hunters tended to encounter varying groups of prey. Yet, an important part of the population did not change its behaviour with experience, which probably contributed to an imperfect matching between predator and prey behaviour, resulting in among individual differences in hunting success. Overall, predators along the flexible-specialist continuum achieved similar success, although flexible hunters were slightly less successful as advanced hunters.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018], while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@Szopa-Comley.Ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@McGhee.etal2013a]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

>>>>

The capacity of predators to match their tactic to the behaviour of their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills and expertise to successfully hunt their prey, which may result in individual foraging specialization. Yet, there are very limited empirical assessments showing links between experience, foraging specialization, and hunting success at the individual level, due to the numerous challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to investigate if and how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. Our results show that predators optimized prey consumption as they gained experience. Prey behaviour was an important mediator of this relationship, resulting in differences among predators in the development of expertise. Throughout experience, 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by swithing between slow and fast speeds. The remaining 32% of predators used a specialized tactic of fast movement throughout the study period. Contrary to our predictions, the predator's strategies (i.e. flexible or specialized foraging) was matched only weakly to the preys' speed, suggesting that changes in hunting behaviour were driven more by repeated experience than the prey's speed. Ultimately, flexible foragers were only slightly less successful than specialist foragers, such that both achieved similar success overall. Our observations suggest that experience drives among individual differences in hunting tactics, which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [cornell2019]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are too complex to analyze mathematically [cornell2019]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@FraserFranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. In a different system, @lymberyetal.2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

@@

Learning from experience is predicted to increase individual predator foraging specialization when prey distribution, behaviour, or movement is predictable through time [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, if predators experience unpredictable encounters with their prey, then learning should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010]. We found that predators in *Dead by Daylight* increased their hunting success with experience, suggesting that predator expertise was honed through extensive practice. While there was no directional increase in either behavioural specialization or flexibility with experience at the population level, individual predators developed their own behavioural trajectories over successive encounters with the prey. This was characterized by a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy, with some predators switching between the two with experience. The behavioural strategies of predators and prey tended to match across experience, suggesting that both were adjusting to each other. Specialized cursorial hunters encountered similar groups of prey, while slower and flexible hunters encountered varying groups of prey. Overall, predators along the flexible-specialist continuum achieved similar success, although flexible hunters were slightly less successful as advanced hunters.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@McGhee.etal2013a]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.
